### 2019-10-20 Manuscript/manuscript.tex
small reference fixes to text

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig1_map.pdf}
\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \citealp{Whitmeyer2007a}). The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age. The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel while the diamonds are additional not-rated results from the Paleomagia database.}}}
\label{fig:Laurentia_map}
\end{figure}

@@

In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications. The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results (Evans et al., this volume). While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland. An `A' rating refers to poles that are judged to be of such high-quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. A `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2013a}. Many of these poles are quite valuable for reconstruction and should not be dismissed from being considered in paleogeography reconstructions. For example, there are rich data associated with intrusive lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and such cooling ages are more difficult to robustly constrain than for eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating. However, while any one of these Grenville poles could be interpreted to suffer from temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although note an alternative view of an allochthonous origin discussed below).

@@

The combined paleomagnetic poles can be used to develop paleogeographic reconstructions for Laurentia. The method typically applied in the Phanerozoic is to develop synthesized pole paths either through fitting splines through the data or calculating binned running means where the Fisher mean of poles within a given interval are calculated \citep{Torsvik2012}. Such an approach is more challenging when there are significant gaps in the record as is the case through the early Neoproterozoic Era and the late Paleoproterozoic into early Mesoproterozoic.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_reconstructions.pdf}
\caption{\small{\textbf{Paleogeogrpahic reconstructions of Laurentia at time intervals through the Proterozoic that are well-constrained by paleomagnetic data. These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} and use this model to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}. Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}.}}}
\label{fig:Laurentia_reconstructions}
\end{figure} 
%ADD MORE RECONSTRUCTIONS. 1 per 100 million years perhaps with indication of which ones are well-constrained by models and

@@

 An exception to the paucity of published continuous paleogeographic models for the Precambrian is the Neoproterozoic model of \cite{Merdith2017a} which is shown in comparison to the constraints for Laurentia in Figure \ref{fig:Laurentia_paleolatitude}. The extent to which the implied position of Laurentia in \cite{Merdith2017a} is consistent with the compiled paleomagnetic constraints can be visualized in Figure \ref{fig:Laurentia_paleolatitude}. As noted above, the development of such models is challenging and the researchers need to balance varying constraints. The focus here will be on the extent to which this model satisfies the available paleomagnetic poles for Laurentia. The model does not honor the Grenville loop (e.g. go to moderately high southerly latitudes ca. 1000 Ma) which is a striking departure from the paleomagnetic record and standard paleogeographic models. Additionally, the implemented plate motion strays from the younger poles of the Keweenawan Track and does not honor the Franklin LIP pole despite its `A' Nordic rating. The Franklin pole is taken to be a key constraint at the Tonian/Cryogenian boundary that provides evidence of the supercontinent Rodinia being equatorial and for the Sturtian glaciation having extended to equatorial latitudes.

@@

\bibliographystyle{gsabull}
\small\bibliography{../../references/allrefs}

>>>>

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig1_map.pdf}
\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \citealp{Whitmeyer2007a} with additions for Greenland based on \citealp{St-Onge2009a}). The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age. The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel while the diamonds are additional not-rated results from the Paleomagia database.}}}
\label{fig:Laurentia_map}
\end{figure}

@@

In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications. The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results (Evans et al., this volume). While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland. An `A' rating refers to poles that are judged to be of such high-quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. A `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a}. Many of these poles are quite valuable for reconstruction and should not be dismissed from being considered in paleogeography reconstructions. For example, there are rich data associated with intrusive lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and such cooling ages are more difficult to robustly constrain than for eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating. However, while any one of these Grenville poles could be interpreted to suffer from temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although note an alternative view of an allochthonous origin discussed below).

@@

The combined paleomagnetic poles can be used to develop paleogeographic reconstructions for Laurentia. The method typically applied in the Phanerozoic is to develop synthesized pole paths either through fitting splines through the data or calculating binned running means where the Fisher mean of poles within a given interval are calculated \citep{Torsvik2012a}. Such an approach is more challenging when there are significant gaps in the record as is the case through the early Neoproterozoic Era and the late Paleoproterozoic into early Mesoproterozoic.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_reconstructions.pdf}
\caption{\small{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic that are well-constrained by paleomagnetic data. These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} and use this model to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}. Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}.}}}
\label{fig:Laurentia_reconstructions}
\end{figure} 
%ADD MORE RECONSTRUCTIONS. 1 per 100 million years perhaps with indication of which ones are well-constrained by models and

@@

 An exception to the paucity of published continuous paleogeographic models for the Precambrian is the Neoproterozoic model of \cite{Merdith2017b} which is shown in comparison to the constraints for Laurentia in Figure \ref{fig:Laurentia_paleolatitude}. The extent to which the implied position of Laurentia in \cite{Merdith2017a} is consistent with the compiled paleomagnetic constraints can be visualized in Figure \ref{fig:Laurentia_paleolatitude}. As noted above, the development of such models is challenging and the researchers need to balance varying constraints. The focus here will be on the extent to which this model satisfies the available paleomagnetic poles for Laurentia. The model does not honor the Grenville loop (e.g. go to moderately high southerly latitudes ca. 1000 Ma) which is a striking departure from the paleomagnetic record and standard paleogeographic models. Additionally, the implemented plate motion strays from the younger poles of the Keweenawan Track and does not honor the Franklin LIP pole despite its `A' Nordic rating. The Franklin pole is taken to be a key constraint at the Tonian/Cryogenian boundary that provides evidence of the supercontinent Rodinia being equatorial and for the Sturtian glaciation having extended to equatorial latitudes.

@@

\bibliographystyle{gsabull}
\footnotesize\bibliography{../../references/allrefs}
